Higher-Order unification for free!

### 2024-04-24 paper/main.tex
minor TODOs

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

@@

\noindent
An example of a \maybebeta{} term is the application $F \appsep x$.
This term is problematic since the application node of 
its syntax tree cannot be used to justify a
unification failure, i.e. by properly instantiating $F$, the term\noteD{or a name}
head constructor may become a $\lambda$, or a constant, or remain an application.

@@

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$\noteD{a,b are reserved for constants}
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms\noteD{module var reflexivity}. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$\noteD{shouldn't be uva Y [] \Ue t} that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
\todo{check progress-rhs is back}
\noteD{first rule is wrong, since maybe-eta is 1-argument pred}

>>>>

\begin{proposition}[Good unification]\label{prop:good-unif}
A good unification $\UnifRel\;$ for equality $\;\EqualRel\;$ in the domain
$\mathcal{D}$ satisfies the following properties:\noteD{use $\sigma$, not $\rho$}
  \begin{gather}
    \unifcorrectX{\EqualRel}{\UnifRel}{t}{\{~t_1,~ t_2~\} \subseteq \mathcal{D}}\\
    \unifcompleteX{\EqualRel}{\UnifRel}{t}{\{~ t_1,~ t_2 ~\} \subseteq \mathcal{D}}
  \end{gather}
\end{proposition}

@@

\noindent
An example of a \maybebeta{} term is the application $F \appsep x$.
This term is problematic since the application node of 
its syntax tree cannot be used to justify a
unification failure, i.e. by properly instantiating $F$, the term
head constructor may become a $\lambda$, or a constant, or a name, or remain an application.

@@

\noindent
An example of a term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda x.\lambda y.f \appsep y \appsep x\}$
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$,
which is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms\noteD{module var reflexivity}. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R, !,
  declare_constraint (link-eta L R) [L,R].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
